---
title: "Package Quick Start Guide"
author: 
- name: Jiefei Wang
  affiliation: Roswell Park Comprehensive Cancer Center, Buffalo, NY
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc: true
        toc_float: true
vignette: >
  %\VignetteIndexEntry{quickStart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
package: AltWrapper
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library("AltWrapper")
```

# Introduction
There are numerous data sources in R. When data is not in a standard R format, R provide a variaty of methods to read it. One can write a function to convert the data into an R data structure, or implement an S3/S4 class and keep the data in its original format. However, this flexibility introduces a lot of troubles when the data is interacting with the other R packages. While one package may assume the data is an R vector, the other package may expect the data is an S4 object with `[` and `[<-` implemented. Without a standard, it will waste a lot of time to handle the data structure, instead of doing the data analysis. 

`ALTREP` is a set of new APIs in R provided since R 3.5. It is designed to represent a non-standard data structure to a standard R vector using the unified APIs. At R level, there is no difference between an R vector and an `ALTREP` vector. This mechanism provides a possibility to wrap any data to a lightweight R vector without using S3/S4 class while still preserving it original data format. While `ALTREP` has the potential to be an official way to read any non-standard data in R, the requirement of knowing C++ language and R's C interfact and lacking of official document impede its development. Even for those who is familar with C++ language, it will takes a lot of energy to learn `ALTREP` and manage their C++ code. `AltWrapper` is a package for developers to make the use of `ALTREP` simpler in R. Developers are able to define all `ALTREP` APIs using R functions, and to debug and replace an `ALTREP` function at any time without doing compilation. While R code is generally slower compared to C++ code,  a special consideration is token to minimize the overhead of calling an user-defined R function. With the help of `AltWrapper`, Users are not required to know C++ language to use `ALTREP`.

# Example

We will illustrate the package with a simple example. Suppose we want to represent a sequence vector in a compact format. The sequence only has a starting value and its increment as the data. We will create a compacted R vecotr by defining `ALTREP` APIS.

```{r}
#sequence 1:10
A_compact = list(start = 1L, by = 1L, length = 10L)

#Define ALTREP APIs
lengthFunc <- function(x) {
  return(x$length)
}
getElementFunc <- function(x, i) {
  return(x$start + x$by * (i - 1))
}

#Register ALTREP class and functions
setAltClass(className = "compactSeq", "integer")
setAltMethod(className = "compactSeq", getLength = lengthFunc)
setAltMethod(className = "compactSeq", getElement = getElementFunc)

#Create altWrapper object
A_vector = makeAltrep(className = "compactSeq", x = A_compact)

#Check the result
is.vector(A_vector)
A_vector[1:10]
```
The return value is an R vector. The value can be accessed via regular `[` operator. Please note that the variable `A_vector` cannot be printed just by typing its variable name in the console since 
ALTREP is still underdevelopment and some R functions do not catch up with it. The problem will be fixed in future R release.














