---
title: "Package Quick Start Guide"
author: 
- name: Jiefei Wang
  affiliation: Roswell Park Comprehensive Cancer Center, Buffalo, NY
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc: true
        toc_float: true
vignette: >
  %\VignetteIndexEntry{quickStart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
package: AltWrapper
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library("AltWrapper")
```

# Introduction
There are numerous data sources in R. When data is not in a standard R format, R provide a variaty of methods to read it. One can write a function to convert the data into an R data structure, or implement an S3/S4 class and keep the data in its original format. However, this flexibility introduces a lot of troubles when the data is interacting with the other R packages. While one package may assume the data is an R vector, the other package may expect the data is an S4 object with `[` and `[<-` implemented. Without a standard, it will waste a lot of time to handle the data structure, instead of doing the data analysis. 

`ALTREP` is a set of new APIs in R provided since R 3.5. It is designed to represent a non-standard data structure to a standard R vector using the unified APIs. At R level, there is no difference between an R vector and an `ALTREP` vector. This mechanism provides a possibility to wrap any data to a lightweight R vector without using S3/S4 class while still preserving it original data format. While `ALTREP` has the potential to be an official way to read any non-standard data in R, the requirement of knowing C++ language and R's C interfact and lacking of official document impede its development. Even for those who is familar with C++ language, it will takes a lot of energy to learn `ALTREP` and manage their C++ code. `AltWrapper` is a package for developers to make the use of `ALTREP` simpler in R. Developers are able to define all `ALTREP` APIs using R functions, and to debug and replace an `ALTREP` function at any time without doing compilation. While R code is generally slower compared to C++ code,  a special consideration is token to minimize the overhead of calling an user-defined R function. With the help of `AltWrapper`, Users are not required to know C++ language to use `ALTREP`.

# Example

We will illustrate the package with a simple example. Suppose we want to represent a sequence vector in a compact format. The sequence only has a starting value and its increment as the data. We will create a compacted R vecotr by defining `ALTREP` APIS.

```{r}
#sequence 1:10
A_compact = list(start = 1L, by = 1L, length = 10L)

#Define ALTREP APIs
lengthFunc <- function(x) {
  return(x$length)
}
getElementFunc <- function(x, i) {
  return(x$start + x$by * (i - 1))
}

## Register ALTREP class and functions
## The type of the vector is integer
setAltClass(className = "compactSeq", "integer")
setAltMethod(className = "compactSeq", getLength = lengthFunc)
setAltMethod(className = "compactSeq", getElement = getElementFunc)

#Create altWrapper object
A_vector = makeAltrep(className = "compactSeq", x = A_compact)

#Check the result
is.vector(A_vector)
```
We claim that the altWrapper class is an integer class. Therefore, the return value of the function `makeAltrep` is an integer R vector. The value can be accessed via regular `[` operator. An altWrapper object must have `getLength` defined. For most operations, altWrapper API `getDataptr` is required. Since there is no actual data for a compact sequence, we define the API `getElement` instead. Please refer to `?setAltMethod` to see a full list of available APIs.

Please note that the variable `A_vector` cannot be printed just by typing its variable name in the console for 
ALTREP is still underdevelopment and some R functions do not catch up with it. The problem will be fixed in future R release. In the current stage, the variable can be printed out either by using `[` operator to force R to use `get_Elt` method, or by using `printAltWrapper` function providing  by AltWrapper package.
```{r}
## Use [ operator
A_vector[1:10]

## Use print method
printAltWrapper(A_vector)
```
Alternatively, there is an S3 print method defined for the altWrapper class, you can create an S3 object by calling the function `makeAltrep` with the argument `S3Class=TRUE`.
```{r}
#Create an S3 version of altWrapper object
A_vector_s3 = makeAltrep(className = "compactSeq", x = A_compact, S3Class=TRUE)
A_vector_s3
#Check class type
class(A_vector)
class(A_vector_s3)
```


# Manage AltWrapper Class
## Function Define/Redefine
The definition of an API of an altWrapper class is ad hoc. You can change the API even after an object is created. For example, we can define the sum function of the class `compactSeq` given that we have created an object `A_vector` for the class. 
```{r}
sumFunc <- function(x, na.rm) {
  message("Sum function")
  a1 = x$start
  n = x$length
  d = x$by
  return(n * a1 + n * (n - 1) * d / 2)
}

setAltMethod(className = "compactSeq", sum = sumFunc)

sum(A_vector)
```
The variable `A_vector` has been made before we define the sum function for its class, but we still see the message from our customized sum function. We can also redefine the sum function at any moment.
```{r}
sumFunc <- function(x, na.rm) {
  message("Sum function V2")
  a1 = x$start
  n = x$length
  d = x$by
  return(n * a1 + n * (n - 1) * d / 2)
}

setAltMethod(className = "compactSeq", sum = sumFunc)

sum(A_vector)
```
A warning will be shown to indicate the sum function has been defined previously. It can be suppressed by calling `setAltWrapperOptions(redefineWarning=FALSE)`. A function can be deleted by setting its value to `NULL`. For example, The sum function can be removed from the class `compactSeq` via `setAltMethod(className = "compactSeq", sum = NULL)`.


## Check Class Functions
The following example shows how to retrieve a function from an altWrapper class.
```{r}
sumFunc = getAltMethod(className = "compactSeq", methodName = "sum")
sumFunc
```
A NULL value will be returned if the function is not found. Note that there is an extra argument `.self` in `sumFunc` function. This argument is used to pass the altWrapper object itself. 

You can also get a summary report of an altWrapper class
```{r}
altClassStatus(className = "compactSeq")
```
The report shows the type of the ALTREP class and status of all ALTREP APIs. The attached function number are the functions which are not ALTREP APIs and associated with the altWrapper class. Please refer to `?attachAltMethod` to see the details.


# Session Info
```{r}
sessionInfo()
```



















